---
title: "Merging and Manipulating ROC plots"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Merging and Manipulating ROC plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
collapse = TRUE,
comment = "#>"
)
```

This vignette will demonstrate:

 * Plotting a ROC curve for a dCVnet object
 * ROC plots with data from different dCVnet models.
 * ROC plot customisation

First, setup the R environment and make some example dCVnet objects:

```{r setup, echo=TRUE, warning=FALSE, message=FALSE, results='hide'}
library(ggplot2)
library(dplyr)
library(dCVnet)

data(prostate, package = "dCVnet")

prostate$agecat <- factor(prostate$age > 65,
                          levels = c(FALSE, TRUE),
                          labels = c("le65", "over65"))

set.seed(42)
# Model 1 predicts SVI (seminal vesicle invasion):
mod1 <- dCVnet::dCVnet(y = prostate$svi,
                       data = subset(prostate,
                                     select = c(-train, -svi)),
                       nrep_outer = 3, k_outer = 3,
                       nrep_inner = 1, k_inner = 10)

# Model 2 predicts whether Age > 65:
mod2 <- dCVnet::dCVnet(y = prostate$agecat,
                       data = subset(prostate, 
                                     select = c(-train, -age, -agecat)),
                       nrep_outer = 3, k_outer = 3,
                       nrep_inner = 1, k_inner = 10)

```

# The default ROC plot
By default, calling `plot(my_dCVnet, type = "ROC")` on a dCVnet object
produces a plot of the cross-validated ROCs for the different outer-loop 
repetitions (to show the variability).

Under the hood this function is composing three steps:

 1. getting performance information from the outer-loop cross-validation
      (using `dCVnet::performance`)
 2. extracting sensitivity, specificity and thresholds and turning this into a 
      standardised *rocdata* format using `dCVnet::extract_rocdata`
 3. running the plot method for rocdata objects (`plot.rocdata`).

```{r default plot, echo=TRUE, fig.height=7, fig.width=7}
p1 <- plot(mod1, type = "roc")

# Note this is the same as:
# p1 <- plot(extract_rocdata(dCVnet::performance(mod1)))

```

# Plots with averaged ROC
To plot overall (average) cross-validated ROC taken over the repetitions of the
repeated k-fold outer-loop cross-validation, use `dCVnet::average_rocdata()` on
the *rocdata* object.

```{r average plot, echo=TRUE, fig.height=7, fig.width=7}
p2 <- plot(average_rocdata(extract_rocdata(performance(mod1))))
# The above nested function is more readable if you use pipes (%>%):
#   p2 <- mod1 %>% 
#           performance %>% 
#           extract_rocdata %>% 
#           average_rocdata %>% 
#           plot()

```

# Combining average and CV-variability
Perhaps we want to see both the average performance, and its variability over 
outerloop-cv repetitions. We can combine two rocdata objects with rbind
(they are just data.frame format with a particular set of columns):

```{r combined plot, echo=TRUE, fig.height=7, fig.width=7}
combined_roc_data <- rbind(extract_rocdata(performance(mod1)),
                           average_rocdata(extract_rocdata(performance(mod1))))
p3 <- plot(combined_roc_data)

# Or, as above, but (mostly) using pipes:
#   p3 <- rbind(mod1 %>% 
#                 performance() %>% 
#                 extract_rocdata(),
#               mod1 %>% 
#                 performance() %>% 
#                 extract_rocdata() %>% 
#                 average_rocdata())


```

# Plotting uncrossvalidated ROC
The 'final model' of dCVnet is the model we cross-validated, it would be used 
to make a prediction for new data. However its performance in the training data 
not crossvalidated, and so is typically an overestimate.

Sometimes we want to see this training set performance all the same - it forms
an upper bound on cross-validated performance. *However, care must always be *
*taken not to interpret this as a cross-validated estimate.*

To plot the uncrossvalidated performance of a final model:
```{r final model, echo=TRUE, fig.height=7, fig.width=7}
plot(extract_rocdata(mod1$final$performance))
```


# Plotting from multiple models
Sometimes we want to display the results from two or more models
in the same ROC plot. Do this by `rbind`-ing the separate rocdata objects.
However, we first need to give informative names in the run column:

```{r multimodel plot, echo=TRUE, fig.height=7, fig.width=7}
d1 <- mod1 %>% 
  performance() %>% 
  extract_rocdata() %>% 
  average_rocdata()
d1$run <- "Model 1" # labels the data from this model.

d2 <- mod2 %>% 
  performance() %>% 
  extract_rocdata() %>% 
  average_rocdata()
d2$run <- "Model 2" # labels the data from this model

d3 <- mod1$final$performance %>% 
  extract_rocdata() %>% 
  average_rocdata()
d3$run <- "Model 1 (train)" # labels the data from this model.

d4 <- mod2$final$performance %>% 
  extract_rocdata() %>% 
  average_rocdata()
d4$run <- "Model 2 (train)" # labels the data from this model.

p4 <- plot(rbind(d1, d2, d3, d4))


```

# Plot options

A good deal of customisation can be done with ggplot (see below), but there are
a few options provided for plot customisation in `plot.rocdata`. The legend can
be toggled off with `legend = FALSE`, different threshold label points can be
chosen (`alphalabel = c(0.05, 0.5, 0.95)`) or these can be omitted
(`alphalabel = FALSE`). Finally alternative labels can be provided for 
legend headings with the `guide_labels` argument.

```{r options1, echo=TRUE, fig.height=7, fig.width=7}
# Make a version of p4 without a legend or markers for threshold:
p5 <- plot(rbind(d1, d2),
           alphalabel = FALSE,
           legend = FALSE)

# And one with custom threshold markers and legend headings:
p5 <- plot(rbind(d1, d2),
           alphalabel = c(0.5),
           guide_labels = list(group = "plotdata",
                               threshold = "50% Threshold",
                               refline = "line"))

```

# Customising plots

dCVnet ROC plots are ggplot2 objects. As a result many manipulations can be done
without recalculating the plot. In the following example I will modify `p4$plot`
(the multiple-models plot) in order to...

 1. Add a plot title and change axis labels
 2. Manually set colours
 3. Change the plot appearance with a different ggplot2 theme
 4. Suppress the legend for the reference line but not other plot features
 
Note: ggplot2 legends are determined by aesthetics which are mapped to the data,
in this this plot Model is mapped to *colour*, Threshold markers to *shape*
and the reference line to *lty* (linetype).

```{r options2, echo=TRUE, fig.height=7, fig.width=7}
p4$plot +
  # add a title, change axis labels:
  labs(title = "Comparison of ROC Performance") +
  xlab("1 - Specificity") +
  ylab("Sensitivity") +
  # set manual colours:
  scale_colour_manual(values = c(`Model 2` = "hotpink4",
                                 `Model 2 (train)` = "hotpink",
                                 `Model 1` = "blue4",
                                 `Model 1 (train)` = "blue")) +
  # suppress part of the legend:
  guides(lty = guide_none()) +
  # use a different theme:
  theme_classic() 

```

# Further Customisation

If you need more control over the display, or would prefer to plot using 
base R, or different software entirely, the underlying data are
returned by the call to `plot.rocdata`.

```{r plotdata, echo=TRUE}
head(p4$data)
```


